### 2022-03-21 manuscript.tex
Update 01.introduction.md [ci skip]
This build is based on
https://github.com/scherzerthesis/thesis/commit/419ff731062cbb3fe3b1423f6c6a6cec111ea028.

This commit was created by the following CI build and job:
https://github.com/scherzerthesis/thesis/commit/419ff731062cbb3fe3b1423f6c6a6cec111ea028/checks
https://github.com/scherzerthesis/thesis/actions/runs/2017477262

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/5c8aea3d5fbc147f384068723bceba76aa35d4a9/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/5c8aea3d5fbc147f384068723bceba76aa35d4a9}{scherzerthesis/thesis@5c8aea3}
on March 18, 2022.

@@

\textbf{MAPK signaling in cancer}

standard activation from the top of he MAPK pathway begins with extracellular ligands (EGF, NGF, etc.) binding to Receptor-Tyrosine Kinases (RTKs). This leads to an loading of RAS with GTP, which activates RAS. Many RAS mutations lead to either more efficient loading of GTP or less effciency GTP hydrolysis which both serve to activate RAS. GTP-loaded RAS recruits and activates RAF kinases (ARAF, BRAF, CRAF) which then phosphorylates and activates MEK1/2 kinases which then phosphorylates and activates ERK1/2. This ultimately leads to the activation of a suite of cellular processes that are required for proliferation, differentiation, migration, and cell survival. Therefore, activating mutation in this pathway, from RTKs to kinases, serve to keep the pathway in a constitutively active state. For example, the most common mutation in \emph{BRAF} is V600E. This valine to glutamic acid substitution is located in the kinase domain and mimics phosphorylation, allowing BRAF(V600E) to become independet of upstream regulations like RAS. This active state drives many of the hallmarks of cancer that is required for tumor formation. It is important to note that oncogenic activation of the MAPK pathway differs from normal ligand-mediated pathway activation. For example, mitogens, such as EGF, only transiently activate the pathway even with constant ligand stimulation. This immediate and strong spike in pathway activation is integrated by various biological processes in the cell that results in transcription of certain target genes. In contrast, oncogenic signaling thru mutational activation of BRAF(V600E) results in constant kinase signaling that is integrated differently than normal pathway stimulation. Thus, an oncogene-specific transcriptional program is activated.

@@

\textbf{BRAF (V600E) signaling and cooperating genetic events to LUAD initiation and progression}

Due to the robust connection between lung adenocarcinomas and mutational activation of genes involved in the MAPK-pathway, such as \emph{EGFR}, \emph{KRAS}, and \emph{BRAF}, many labs have generated genetically-engineered mouse models (GEMMs) that harbor conditionally-activated cancer specific mutant alleles of either EGFR (L858R), KRAS (G12D), or BRAF(V600E). Expression of BRAF (V600E) in the Surfactant-Protein C expression cells (alveolar type 2 pneumocytes) elicits benign tumor formation that fail to progress to malignant lung adenocarcinoma. It is hypothesized that the cell cycle arrest is dependent on the tumor suppressors P53, and CDKN2A, as loss of either P53 or CDKN2A allows tumors to progress to malignant and deadly lung adenocarcinoma. Also, activation of WNT signaling and PI3K signaling similarly allows BRAF (V600E) adenomas to progress to adenocarcinoma. These observations are in concordance to the "multiple-hit" hypothesis. This hypothesis states that cancers do not arise from single mutagenic events, but are the consequence of sequential assaults on the genome that activate more than one oncogenic pathway. These multiple genetic hits allow cells to adopt traits to form cancer, which is coined by Bob Weinberg and Douglas Hanahan as the "Hallmarks of Cancer". The traits include suppression of apoptosis, active proliferation, altered metabolism, changes in cell identity, and forming blood-vessels (Hanahan, 2022). BRAF(V600E) expression alone is sufficient to activate many of the hallmarks of cancer and additional pathways need to be activated to form deadly metastasis.

@@

\textbf{Genetically engineered mouse models (GEMMs)}

Genetically-Engineered mouse models (GEMMs) of human cancer have allowed for the reproducible analysis of cancer biology driven by specific mutations that are often found in cancer patients. These models are useful for understanding the molecular mechanisms that drive tumor formation but also serve as a platform for testing tumor responses to targeted- or immune- therapies. Early mouse models utilized either the inherent tumor susceptibility of inbred strains of mice, such as FVB , or carcinogen-induced models such as UV, Urethane, or DMBA treatment. With the ability to genetically manipulate embryonic stem-cells (ES cells) came the development of mice with mutations engineered to spontaneously form neoplasms without the need for carcinogen treatments. Genetic knockouts or over-expression can lead to embryonic lethality depending upon the specific role of the manipulated gene, therefore conditional gene manipulation was developed to allow for temporal control thus overcoming the issues with embryonic lethality or other developmental defects.

The most common conditional system relies on chemically inducible transcription factors, such as tetracycline-dependent regulatory system to relies on the inoculation of tetracycline into the body to bind to a tetracycline-trans-activator to turn on gene expression. More recently, Cre-Lox recombination strategies have been utilized to get more cell type specific control. In this system, genes of interested are constructed to contain LoxP-sites that flank a particular genetic element and upon delivery of Cre-recombinase the DNA sequence in between LoxP sites are excised. With this approach, one can knockout a gene by engineering LoxP sites around critical exons, or one could turn on gene transcription by flanking a strong stop signal such as a polyadenylation sequence upstream of a gene-of-interest sequence such as KRAS G12D.

@@

Our lab has previously developed a conditional mouse model of BRAF (V600E) human lung cancers. In the \emph{Braf(CAT)} model, normal BRAF is expressed from a conditional allele prior to Cre-mediated recombination . LoxP sites flank human \emph{BRAF} cDNA encoding normal \emph{BRAF} exons 15-18. Downstream of the LoxP sites is the mutant exon 15 that encodes the murine equivalent of the T1799A mutation that gives rise to the BRAF (V600E) onco-protein. Downstream of the mutant exon there is a P2A element and a CAAX-tagged \emph{TdTomato}. Therefore, after Cre-mediated recombination, the BRAF (V600E) oncoprotein and fluorophore TdTomato is expressed at normal levels.

@@

\textbf{P53- mediated tumor suppression}

In biology, there are highly conserved pathways that control a bewildering amount of cellular processes that are both dependent and independent of each other, such as NOTCH- or WNT- signaling. The same could be said about P53. At the time of writing this dissertation, there are approximately 105,000 manuscripts on PUBMED that mention P53. P53 is so well studied because of both its apparent role in tumor suppression, but also because it is the most frequently altered gene in human cancer.

P53 is a transcriptional factor that has a DNA-binding domain that specifically recognizes two decameric half-sites. P53 also contains two N-terminal transactivation domains and a C-terminal oligomerization domain that are critical for P53-target gene activation. Central to the ability of P53 to induce transcriptional activation is tetramerization. Thus, if one or more P53 proteins in the complex is compromised in transactivation or DNA binding, transcriptional activity is compromised. Therefore, if one copy of P53 is mutated, the organism's ability to suppress tumor formation is compromised.

The importance of P53 is best demonstrated in people, and mice, that have just one mutated copy of P53. In humans with the familial- inherited Li-Fraumeni syndrome, in which they have one or more mutations in p53 are almost certain to develop cancer early on and throughout their life. Similarly, mice lacking two function P53 genes are prone to leukemias and lymphomas and have a shortened lifespan due to cancer incidence.

The most well-known cellular functions of P53 include its ability to induce cell-cycle arrest in response to DNA damage. Many stress signals, including oncogene activation has been shown to stimulate a reversible or irreversible cell cycle arrest. However, the strength of P53 induced cell cycle arrest or apoptosis is likely cell type- and cellular stress- specific.

@@

P53 was originally thought to be an oncogene as P53 is accumulated in many human tumors which is not common in normal tissues. Moreover, ectopic expression of a P53 cDNA was found to aide in the transformation of primary cells induced by RAS. However, early studies erroneously used mutated P53 instead of wild-type leading to the misclassification of P53 as an oncogene. We known know P53 suppresses cell growth and transformation.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/419ff731062cbb3fe3b1423f6c6a6cec111ea028/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/419ff731062cbb3fe3b1423f6c6a6cec111ea028}{scherzerthesis/thesis@419ff73}
on March 21, 2022.

@@

\textbf{MAPK signaling in cancer} standard activation from the top of he MAPK pathway begins with extracellular ligands (EGF, NGF, etc.) binding to Receptor-Tyrosine Kinases (RTKs). This leads to an loading of RAS with GTP, which activates RAS. Many RAS mutations lead to either more efficient loading of GTP or less effciency GTP hydrolysis which both serve to activate RAS. GTP-loaded RAS recruits and activates RAF kinases (ARAF, BRAF, CRAF) which then phosphorylates and activates MEK1/2 kinases which then phosphorylates and activates ERK1/2. This ultimately leads to the activation of a suite of cellular processes that are required for proliferation, differentiation, migration, and cell survival. Therefore, activating mutation in this pathway, from RTKs to kinases, serve to keep the pathway in a constitutively active state. For example, the most common mutation in \emph{BRAF} is V600E. This valine to glutamic acid substitution is located in the kinase domain and mimics phosphorylation, allowing BRAF(V600E) to become independet of upstream regulations like RAS. This active state drives many of the hallmarks of cancer that is required for tumor formation. It is important to note that oncogenic activation of the MAPK pathway differs from normal ligand-mediated pathway activation. For example, mitogens, such as EGF, only transiently activate the pathway even with constant ligand stimulation. This immediate and strong spike in pathway activation is integrated by various biological processes in the cell that results in transcription of certain target genes. In contrast, oncogenic signaling thru mutational activation of BRAF(V600E) results in constant kinase signaling that is integrated differently than normal pathway stimulation. Thus, an oncogene-specific transcriptional program is activated.

@@

@@

\textbf{Genetically-engineered mouse models }Genetically-Engineered mouse models (GEMMs) of human cancer have allowed for the reproducible analysis of cancer biology driven by specific mutations frequently found in cancer patients. These models are helpful in understanding the molecular mechanisms that drive tumor formation and testing tumor responses to targeted or immune therapies. Early mouse models were not engineered but instead selected inbred mouse strains that were natural hosts of tumor transplantation(Strong, 1978). However, these models suffer from long latencies, incomplete pentrance, and often not relevant to human cancers. Therefore, carcinogen-induced models such as UV, Urethane, radiation, cigarette consumption, or MNU treatment models were developed to ressemble carinogene-induced human cancers(Kemp \emph{et al.}, 1994; Westcott \emph{et al.}, 2014).After techniques were developed to genetically manipulate embryonic stem-cells came the ability to engineer mice with specific mutations that could drive tumor formation. Since genes involved in cancer are also important for normal mouse development, conditional gene manipulation was developed to allow for temporal control.

The most commonly used conditional gene expression system relies on inducible transcription factors, such as tetracycline-dependent regulatory system, where the treatment of mice with doxycycline toggles gene transcriptial levels of transgenes. More recently, Cre-Lox recombination strategies have been utilized to get more cell type specific control. In this system, genes of interested are constructed to contain LoxP-sites that flank a particular genetic element and upon delivery of Cre-recombinase the DNA sequence in between LoxP sites are excised. With this approach, one can knockout a gene by engineering LoxP sites around critical exons, or one could induce gene transcription by flanking a strong stop signal such as a polyadenylation sequence upstream of a gene-of-interest sequence such as KRAS(G12D)(Jackson \emph{et al.}, 2001).

@@

Due to the robust connection between lung adenocarcinomas and mutational activation of genes involved in the MAPK-pathway, such as \emph{EGFR}, \emph{KRAS}, and \emph{BRAF}, many labs have generated genetically-engineered mouse models (GEMMs) that harbor conditionally-activated cancer specific mutant alleles of either EGFR (L858R), KRAS (G12D), or BRAF(V600E).

Our lab has previously developed a conditional mouse model of BRAF (V600E) human lung cancers(Dankort \emph{et al.}, 2007; Shai \emph{et al.}, 2015; van Veen \emph{et al.}, 2019). In the \emph{Braf(CAT)} model, normal BRAF is expressed from a conditional allele prior to Cre-mediated recombination . LoxP sites flank human \emph{BRAF} cDNA encoding normal \emph{BRAF} exons 15-18. Downstream of the LoxP sites is the mutant exon 15 that encodes the murine equivalent of the T1799A mutation that gives rise to the BRAF (V600E) onco-protein. Downstream of the mutant exon there is a P2A element and a CAAX-tagged \emph{TdTomato}. Therefore, after Cre-mediated recombination, the BRAF (V600E) oncoprotein and fluorophore TdTomato is expressed at normal levels(van Veen \emph{et al.}, 2019).

\textbf{BRAF (V600E) signaling and cooperating genetic events to LUAD initiation and progression} Mutational activation of the MAPK pathway forms lung tumors in mice. specifically, expression of BRAF (V600E) in the Surfactant-Protein C expression cells (alveolar type 2 pneumocytes) elicits benign tumor formation that fails to progress to lung adenocarcinoma. It is hypothesized that the cell cycle arrest is dependent on the tumor suppressors P53 and CDKN2A. Loss of either P53 or CDKN2A allows tumors to progress to malignant and deadly lung adenocarcinoma. Also, activation of WNT signaling and PI3K signaling similarly allows BRAF (V600E) adenomas to progress to adenocarcinoma (van Veen \emph{et al.}, 2019; Green \emph{et al.}, 2015; Juan \emph{et al.}, 2014). These observations are in concordance to the "multiple-hit" hypothesis that states that cancers do not arise from single mutagenic events, but are the consequence of sequential assaults on the genome that activate more than one oncogenic pathway. These multiple genetic hits allow cells to adopt traits to form cancer, which is coined by Bob Weinberg and Douglas Hanahan as the "Hallmarks of Cancer" (Hanahan and Weinberg, 2011; Hanahan, 2022). The traits include suppression of apoptosis, active proliferation, altered metabolism, changes in cell identity, and forming blood-vessels to supply tumor cells with adequate resources for growth(Hanahan, 2022). BRAF(V600E) expression alone is sufficient to activate many of the hallmarks of cancer but additional pathways need to be activated, or suppresed, to develop to malignant tumors. Great effort has elucidated the proto-oncogenes that drive cancer formation when mutated, such as \emph{MYC}, \emph{RAS}, \emph{PI3K}, and others. Yet we still do not fully understand the mechanistic details that determine how genes cooperate to overide tumor suppressive mechanisms and drive cancer progression.

@@

\textbf{P53- mediated tumor suppression}
Cancers pose an evoltionary challenge for organisms regardless of their life span. Therefore, supression of the outgrowth of neoplastic cells is built into many organisms. One such mechanism is the engagment of the p53 pathway. At the time of writing this dissertation, there are approximately 105,000 manuscripts on PUBMED that mention P53. P53 is so well studied because of both its apparent role in tumor suppression, but also because it is the most frequently altered gene in human cancer.

The p53 protein was originally discovered as a protein that bound large T antigene of the SV40 virus in infected cells (Linzer and Levine, 1979).P53 is a transcriptional factor that has a DNA-binding domain that specifically recognizes two decameric half-sites. P53 also contains two N-terminal transactivation domains and a C-terminal oligomerization domain that are critical for P53-target gene activation(Laptenko and Prives, 2006). Central to the ability of P53 to induce transcriptional activation is tetramerization. Thus, if one or more P53 proteins in the complex is compromised in transactivation or DNA binding, transcriptional activity is compromised. Therefore, if one copy of P53 is mutated, the organism's ability to suppress tumor formation is compromised.

The importance of p53 is best demonstrated in people, and in mice, that have just one mutated copy of P53. In humans with the familial- inherited Li-Fraumeni syndrome, in which they have one or more mutations in p53 are almost certain to develop cancer early on and throughout their life(Li \emph{et al.}, 1988). Similarly, mice lacking two function P53 genes are prone to leukemias and lymphomas and have a shortened lifespan due to cancer incidence(Donehower \emph{et al.}, 1992). On the other hand, organisms such as the african and asian elephants, which have an abnormally low level of cancer incidence, have been shown to higher multiple copies of \emph{TP53} in their genomes(Abegglen \emph{et al.}, 2015).

The most well-known cellular functions of P53 include its ability to induce cell-cycle arrest in response to DNA damage(el-Deiry \emph{et al.}, 1993). Many stress signals, including oncogene activation has been shown to stimulate a reversible or irreversible cell cycle arrest(Zhu \emph{et al.}, 1998; McNeal \emph{et al.}, 2021). However, the strength of P53 induced cell cycle arrest or apoptosis is likely cell type- and cellular stress- specific and is a active area of p53 research.

@@

P53 was originally thought to be an oncogene as P53 is accumulated in lung cancer, which is not common in normal tissues(Iggo \emph{et al.}, 1990; Bartek \emph{et al.}, 1990). Moreover, ectopic expression of a P53 cDNA was found to aide in the transformation of primary cells induced by RAS. However, early studies erroneously used mutated P53 instead of wild-type leading to the misclassification of P53 as an oncogene. We now know P53 suppresses cell growth and transformation(Levine and Oren, 2009).

@@

\leavevmode\vadjust pre{\hypertarget{ref-JyurLVYi}{}}%
Abegglen,L.M. \emph{et al.} (2015) \href{https://doi.org/10.1001/jama.2015.13134}{Potential Mechanisms for Cancer Resistance in Elephants and Comparative Cellular Response to DNA Damage in Humans.} \emph{JAMA}, \textbf{314}, 1850--60.

@@

\leavevmode\vadjust pre{\hypertarget{ref-HnQw16l7}{}}%
Bartek,J. \emph{et al.} (1990) \href{https://www.ncbi.nlm.nih.gov/pubmed/1694291}{Genetic and immunochemical analysis of mutant p53 in human breast cancer cell lines.} \emph{Oncogene}, \textbf{5}, 893--9.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kxXMFjnc}{}}%
Donehower,L.A. \emph{et al.} (1992) \href{https://doi.org/10.1038/356215a0}{Mice deficient for p53 are developmentally normal but susceptible to spontaneous tumours.} \emph{Nature}, \textbf{356}, 215--21.

\leavevmode\vadjust pre{\hypertarget{ref-IILnKdOc}{}}%
el-Deiry,W.S. \emph{et al.} (1993) \href{https://doi.org/10.1016/0092-8674(93)90500-p}{WAF1, a potential mediator of p53 tumor suppression.} \emph{Cell}, \textbf{75}, 817--25.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Pje1Soz4}{}}%
Green,S. \emph{et al.} (2015) \href{https://doi.org/10.1158/0008-5472.can-15-1249}{PIK3CA(H1047R) Accelerates and Enhances KRAS(G12D)-Driven Lung Tumorigenesis.} \emph{Cancer Res}, \textbf{75}, 5378--91.

@@

\leavevmode\vadjust pre{\hypertarget{ref-4H4Kjv9w}{}}%
Hanahan,D. and Weinberg,R.A. (2011) \href{https://doi.org/10.1016/j.cell.2011.02.013}{Hallmarks of cancer: the next generation.} \emph{Cell}, \textbf{144}, 646--74.

@@

\leavevmode\vadjust pre{\hypertarget{ref-bLXbRmvx}{}}%
Iggo,R. \emph{et al.} (1990) \href{https://doi.org/10.1016/0140-6736(90)90801-b}{Increased expression of mutant forms of p53 oncogene in primary lung cancer.} \emph{Lancet}, \textbf{335}, 675--9.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Djv3zv5n}{}}%
Juan,J. \emph{et al.} (2014) \href{https://doi.org/10.1101/gad.233627.113}{Diminished WNT -\textgreater{} β-catenin -\textgreater{} c-MYC signaling is a barrier for malignant progression of BRAFV600E-induced lung tumors.} \emph{Genes Dev}, \textbf{28}, 561--75.

@@

\leavevmode\vadjust pre{\hypertarget{ref-9P0mUQ2t}{}}%
Kemp,C.J. \emph{et al.} (1994) \href{https://doi.org/10.1038/ng0994-66}{p53-deficient mice are extremely susceptible to radiation-induced tumorigenesis.} \emph{Nat Genet}, \textbf{8}, 66--9.

@@

\leavevmode\vadjust pre{\hypertarget{ref-qbfESpXa}{}}%
Laptenko,O. and Prives,C. (2006) \href{https://doi.org/10.1038/sj.cdd.4401916}{Transcriptional regulation by p53: one protein, many possibilities.} \emph{Cell Death Differ}, \textbf{13}, 951--61.

@@

\leavevmode\vadjust pre{\hypertarget{ref-SWNWMYmm}{}}%
Levine,A.J. and Oren,M. (2009) \href{https://doi.org/10.1038/nrc2723}{The first 30 years of p53: growing ever more complex.} \emph{Nat Rev Cancer}, \textbf{9}, 749--58.

\leavevmode\vadjust pre{\hypertarget{ref-TYj65wKz}{}}%
Li,F.P. \emph{et al.} (1988) \href{https://www.ncbi.nlm.nih.gov/pubmed/3409256}{A cancer family syndrome in twenty-four kindreds.} \emph{Cancer Res}, \textbf{48}, 5358--62.

\leavevmode\vadjust pre{\hypertarget{ref-1FKCm8fuC}{}}%
Linzer,D.I. and Levine,A.J. (1979) \href{https://doi.org/10.1016/0092-8674(79)90293-9}{Characterization of a 54K dalton cellular SV40 tumor antigen present in SV40-transformed cells and uninfected embryonal carcinoma cells.} \emph{Cell}, \textbf{17}, 43--52.

@@

\leavevmode\vadjust pre{\hypertarget{ref-1DxvFFUzE}{}}%
McNeal,A.S. \emph{et al.} (2021) \href{https://doi.org/10.7554/elife.70385}{BRAF}. \emph{Elife}, \textbf{10}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-4BJVCLfM}{}}%
Strong,L.C. (1978) \href{https://doi.org/10.1002/jso.2930100410}{Three mice, two blind.} \emph{J Surg Oncol}, \textbf{10}, 353--60.

@@

\leavevmode\vadjust pre{\hypertarget{ref-zOzqGxQM}{}}%
Westcott,P.M.K. \emph{et al.} (2014) \href{https://doi.org/10.1038/nature13898}{The mutational landscapes of genetic and chemical models of Kras-driven lung cancer.} \emph{Nature}, \textbf{517}, 489--92.

@@

\leavevmode\vadjust pre{\hypertarget{ref-N2yNjRyX}{}}%
Zhu,J. \emph{et al.} (1998) \href{https://doi.org/10.1101/gad.12.19.2997}{Senescence of human fibroblasts induced by oncogenic Raf.} \emph{Genes Dev}, \textbf{12}, 2997--3007.
